---
title: "Cancer Prediction"
author: "Group 2 (Colin Fitzpatrick, Sam Mosiman, Brandon Winder)"
date: "10/22/2020"
output: word_document
---
```{r include=FALSE}
# Loads all necessary libraries.
library(ggplot2)
library(dplyr)
library(corrplot)
library(pROC)
library(caret)
```


# Introduction
The goal of our work is to use data on tissue samples taken from cancerous breast tissue to create a logistic regression model that can be used to accurately predict if a cancer sample is malignant or benign.  Our approach is to first analyze all 10 mean predictor variables to check for extreme values that could influence our models accuracy.  We chose to use mean predictors exclusively because the worst predictors should be included in the mean, but we didn't want to focus on the outliers for each specific observation.  Additionally we did not include the standard error variables to predict whether cancer cells are malignant or benign because they are solely based off the mean variables and would not provide much information on the diagnosis.  To do this, we will plot predictor variables and look for outliers that could skew the variable.  If need be, we can then transform the skewed variables through the use of square roots, logarithms, and reciprocals.  We will also look at the predictor variables to determine if any are highly correlated so that we can  reduce the number of predictor variables to account for overfitting our regression model.  Once we have checked our predictor variables and transformed those where we felt necessary, we will build models and test their predictive capabilities.

# Exploring and Transforming the Data
The descriptive statistics for each variable mean of interest are displayed below: 

```{r echo=FALSE}
# Read in the data file and make diagnosis a factor variable.
breastcancer <- read.csv("breastcancer.csv")
breastcancer$diagnosis <-  ifelse(breastcancer$diagnosis == "M", 1, 0)
breastcancer$diagnosis <- as.factor(breastcancer$diagnosis)

# Print the summaries for all variable means
summaries <- as.data.frame(apply(breastcancer[3:12], 2, summary))
round(summaries, digits = 2)
```

```{r fig.dim=c(3,3), include=FALSE}
# First look at the scatter plots of the means.
area_plot = plot(breastcancer$area_mean, breastcancer$diagnosis, xlab="Area Mean", ylab = "Diagnosis", main = "Area Mean against Diagnosis", cex.main = 0.75)

radius_plot = plot(breastcancer$radius_mean, breastcancer$diagnosis, xlab="Radius Mean", ylab = "Diagnosis", main = "Radius Mean against Diagnosis", cex.main = 0.75)

texture_plot = plot(breastcancer$texture_mean, breastcancer$diagnosis, xlab="Texture Mean", ylab = "Diagnosis", main = "Texture Mean against Diagnosis", cex.main = 0.75)

permieter_plot = plot(breastcancer$perimeter_mean, breastcancer$diagnosis, xlab="Perimeter Mean", ylab = "Diagnosis", main = "Perimeter Mean against Diagnosis", cex.main = 0.75)

compact_plot = plot(breastcancer$compactness_mean, breastcancer$diagnosis, xlab="Compactness Mean", ylab = "Diagnosis", main = "Compactness Mean against Diagnosis", cex.main = 0.75)

fractal_plot = plot(breastcancer$fractal_dimension_mean, breastcancer$diagnosis, xlab="Fractal Dimension Mean", ylab = "Diagnosis", main = "Fractal Dimension Mean against Diagnosis", cex.main = 0.75)

concave_point_plot = plot(breastcancer$concave.points_mean, breastcancer$diagnosis, xlab="Concave Points Mean", ylab = "Diagnosis", main = "Concave Points Mean against Diagnosis", cex.main = 0.75)

concavity_plot = plot(breastcancer$concavity_mean, breastcancer$diagnosis, xlab="Concavity Mean", ylab = "Diagnosis", main = "Concavity Mean against Diagnosis", cex.main = 0.75)

symmetry_plot = plot(breastcancer$symmetry_mean, breastcancer$diagnosis, xlab="Symmetry Mean", ylab = "Diagnosis", main = "Symmetry Mean against Diagnosis", cex.main = 0.75)

smooth_plot = plot(breastcancer$smoothness_mean, breastcancer$diagnosis, xlab="Smoothness Mean", ylab = "Diagnosis", main = "Smoothness Mean against Diagnosis", cex.main = 0.75)
```

After analyzing the summaries and the distribution of the variables, we have decided to perform many transformations to account for abnormally distributed data. The transformations on each variable can be seen in the boxplots displayed below.  These transformations were performed because the normal boxplots of the predictors had outliers and/or were highly skewed. Then the data was transformed either by a log, reciprocal, square root, and cube root transformations.  These transformations make the distributions less skewed and contain less outliers, making sure the data is as close to normal as possible.  After performing the transformations, the box plots shown below show less outliers and are less skewed than the originals.  These transformations have a more likely chance of getting better prediction models later when we conduct tests of the data.  Models will also be more accurate with these changes in transformations as there are less outliers to skew the models that we will conduct later.  Overall, these transformations were needed to better fit models and get the data as close to normal as possible. The boxplots shown below help us identify which variables need to be transformed.  When we plot these we are looking for three things in these plots.  We are looking to see how many outliers the data has, the spread of the data and if it is consistent throughtout all of plot, and we are also looking to see if the data is heavily skewed in these boxplots as well.  These three factors will help us determine if we need to make the data fit better to a normal curve.

```{r fig.dim=c(3,3), include=FALSE}
# Display the side-by-side boxplot for malignant and benign cancer samples.
breastcancer %>% ggplot(aes(log(area_mean), diagnosis)) + 
  geom_boxplot() + 
  xlab("Log Area Mean") +
  ylab("Diagnosis") +
  ggtitle("Log Area Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(log(radius_mean), diagnosis)) +
  geom_boxplot() + 
  xlab("Log Radius Mean") +
  ylab("Diagnosis") +
  ggtitle("Log Radius Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(log(texture_mean), diagnosis)) +
  geom_boxplot() + 
  xlab("Log Texture Mean") +
  ylab("Diagnosis") +
  ggtitle("Log Texture Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(log(perimeter_mean), diagnosis)) + 
  geom_boxplot() + 
  xlab("Log Perimeter Mean") +
  ylab("Diagnosis") +
  ggtitle("Log Perimeter Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(log(compactness_mean), diagnosis)) + 
  geom_boxplot() + 
  xlab("Log Compactness Mean") +
  ylab("Diagnosis") +
  ggtitle("Log Compactness Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(1/(fractal_dimension_mean), diagnosis)) + 
  geom_boxplot() + 
  xlab("Reciprical of Fractal Dimension Mean") +
  ylab("Diagnosis") +
  ggtitle("Reciprical Fractal Dimension Mean by Diagnosis") +
  theme(plot.title = element_text(size = 7))

breastcancer %>% ggplot(aes(sqrt(concave.points_mean), diagnosis)) + 
  geom_boxplot() + 
  xlab("Square Root Concavity Points Mean") +
  ylab("Diagnosis") +
  ggtitle("Square Root Concave Points Mean by Diagnosis") +
  theme(plot.title = element_text(size = 7))

breastcancer %>% ggplot(aes((concavity_mean)^(1/3), diagnosis)) +
  geom_boxplot() + 
  xlab("Cube Root Concavity Mean") +
  ylab("Diagnosis") +
  ggtitle("Cube Root Concavity Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(log(symmetry_mean), diagnosis)) + 
  geom_boxplot() + 
  xlab("Log Symmetry Mean") +
  ylab("Diagnosis") +
  ggtitle("Log Symmetry Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

breastcancer %>% ggplot(aes(smoothness_mean, diagnosis)) + 
  geom_boxplot() + 
  xlab("Smoothness Mean") +
  ylab("Diagnosis") +
  ggtitle("Smoothness Mean by Diagnosis") +
  theme(plot.title = element_text(size = 10))

# Modify the data frame with all the transformations.
breastcancer <- breastcancer %>% mutate(area_mean = log(area_mean), radius_mean = log(radius_mean), texture_mean = log(texture_mean), perimeter_mean = log(perimeter_mean), compactness_mean = log(compactness_mean), fractal_dimension_mean = 1/(fractal_dimension_mean), concave.points_mean = sqrt(concave.points_mean), concavity_mean = (concavity_mean)^(1/3), symmetry_mean = log(symmetry_mean))
```

Based on this correlation plot, we can see that the variable 'concave.points_mean' and 'concavity_mean' have a high positive correlation with most other predictor variable means. Also when looking at the predictors standard errors and worst values we see the same trends. So, we will want to pay close attention to these variables when creating the logistic regression model to avoid multicollinearity. Additionally, we will want to examine the variables 'radius_mean', 'perimeter_mean', and 'area_mean' because they are also highly positively correlated. 

```{r echo=FALSE, fig.dim=c(4,4)}
# Create correlations for the predictor variables.
breastcancer_corr <- cor(breastcancer[3:12], use = "complete.obs")

# Create a correlation plot for the predictor variables.
corrplot(breastcancer_corr, type = "upper", tl.cex = 0.5)
```

```{r include=FALSE}
# Find the correlations for the standard errors and worst values.
breastcancer_corr_se <- cor(breastcancer[13:22], use = "complete.obs")
breastcancer_corr_worst <- cor(breastcancer[23:32], use = "complete.obs")

# Find the plots for the correlations above.
corrplot(breastcancer_corr_se, type = "upper", tl.cex = 0.5)
corrplot(breastcancer_corr_worst, type = "upper", tl.cex = 0.5)
```


## The Logistic Model
```{r}
# Creating data frame with only variables of interest
breastcancer1 <- breastcancer[,-c(1,13:32)]

# Partitioning the data into training and validation sets
set.seed(12)
train.index <- createDataPartition(breastcancer1$diagnosis, p = 0.75,
                                  list = FALSE,
                                  times = 1)
train.df = breastcancer1[train.index,]
valid.df = breastcancer1[-train.index,]

# Building model with all predictors
modALL <- glm(diagnosis ~ ., family = binomial, data = train.df)
```

```{r}
# Find the predicted class of the validation data and create contingency table.
pred <- predict(modALL, valid.df, type = "response")
class <- ifelse(pred >= 0.5, 1, 0)
table <- table(class, valid.df$diagnosis)

# Find the sensitivity, specificity, and overall accuracy.
paste("Senstivity: ")
table[2,2]/sum(table[,2])
paste("Specificity: ")
table[1,1]/sum(table[,1])
paste("Overall accuracy: ")
(table[1,1] + table[2,2]) / length(class)
```

## Model Selection
```{r}
# Back step
back.elim <- step(modALL, direction = "backward", trace = 0)
summary(back.elim)

# Forward step
null <- glm(diagnosis ~ 1, family = binomial, data = train.df)
forward.elim <- step(null, scope = list(lower = null, upper = modALL), direction = "forward", trace = 0)
summary(forward.elim)
```

### Back
```{r}
# Find the predicted class of the validation data and create contingency table.
pred2 <- predict(back.elim, valid.df, type = "response")
class2 <- ifelse(pred2 >= 0.5, 1, 0)
table2 <- table(class2, valid.df$diagnosis)

# Find the sensitivity, specificity, and overall accuracy.
paste("Senstivity: ")
table2[2,2]/sum(table2[,2])
paste("Specificity: ")
table2[1,1]/sum(table2[,1])
paste("Overall accuracy: ")
(table2[1,1] + table2[2,2]) / length(class2)
```

### Forward
```{r}
# Find the predicted class of the validation data and create contingency table.
pred3 <- predict(forward.elim, valid.df, type = "response")
class3 <- ifelse(pred3 >= 0.5, 1, 0)
table3 <- table(class3, valid.df$diagnosis)

# Find the sensitivity, specificity, and overall accuracy.
paste("Senstivity: ")
table3[2,2]/sum(table3[,2])
paste("Specificity: ")
table3[1,1]/sum(table3[,1])
paste("Overall accuracy: ")
(table3[1,1] + table3[2,2]) / length(class3)
```

## Optimal Threshold
```{r}
rocfit <- roc(valid.df$diagnosis, class2)
metrics <- rocfit$sensitivities + rocfit$specificities
max.metric.ind <- which(metrics == max(metrics))
opt.threshold <- rocfit$thresholds[max.metric.ind]
opt.threshold
```

